The role of prenatal conditions on brain mitochondrial physiology and prey detection in a lizard

### 2025-04-28 ms.qmd
updates

>>>>

#### What did average mitochondria do?

When obtaining data using FlowJo, we got the geometric mean value for each event recorded in the flow cytometer after gatting for positive values in the dyes employed (see _Methods: Flow cytometry_). However, this is not informative about the effects per mitochondria. In other words, we might be loosing some resolution about the effects at a mitochondrial level rather than a cellular level. To further investigate the effects per mitochondrial unit, we repeated the models above but dividing each variable by mean mitochondrial density obtained for each individual.

```{r, data_mitochondrial_controlled}
#| label: data_mitochondrial_controlled
# 
mit_controlled <- mit_df %>%
    mutate(Id_flow = as.character(Id_flow)) %>%
    mutate(mean_dnadamage = log10(mean_dnadamage + 1)) %>%
    mutate(mean_peroxidation = log10(mean_peroxidation + 1)) %>%
    mutate(across(where(~ is.numeric(.)), ~.x / mean_mitodensity))
#
df_controlled <- merge(data_discr, mit_controlled, by = c("stimulus", "Id_flow")) %>%
  mutate(across(c(age_trial, age_euthanasia), as.numeric)) %>%
  mutate(across(where(~ is.double(.) & !is.integer(.)),
                ~ (. - mean(., na.rm = TRUE)) / (2 * sd(., na.rm = TRUE))))
```

```{r, models_mitochondrial_controlled}
#| label: models_mitochondrial_controlled
#
# Run models mitochondrial physiology from df_controlled (each region separately)
#
var_m_controlled <- c("mean_potential", "mean_ros", "mean_dnadamage", "mean_peroxidation")
regions <- c("OB", "OT")
for (p in var_m_controlled){
  for (h in regions){
    if (h == "OB"){
      df <- df_controlled %>% filter(region == "OB")
      l <- "OB"
      if (p %in% c("mean_mitodensity", "mean_potential", "mean_ros")){
        formula <- paste0(p, "~ cort*temp + (1|clutch)")
      } else if (p == "mean_dnadamage"){
        formula <- paste0(p, "~ cort*temp + age_euthanasia + sex + (1|clutch)")
      } else {
        formula <- paste0(p, "~ cort*temp + age_euthanasia + (1|clutch)")
      }
    } else {
      df <- df_controlled %>% filter(region == "OT")
      l <- "OT"
      if (p %in% c("mean_mitodensity", "mean_potential", "mean_ros")){
        formula <- paste0(p, "~ cort*temp + (1|clutch)")
      } else if (p == "mean_dnadamage"){
        formula <- paste0(p, "~ cort*temp + sex + (1|clutch)")
      } else {
        formula <- paste0(p, "~ cort*temp + age_euthanasia + (1|clutch)")
      }
    }

    pmodel_name <- paste0("m_controlled_", p, "_", h)
    assign(pmodel_name, fit_m(df = df,
                              cat = "controlled",
                              var = p,
                              formula = formula,
                              fam = gaussian(),
                              label = l,
                              refit = FALSE),
          envir = .GlobalEnv)  # Assign to the global environment
  }
} 
```

```{r, organise_posteriors_controlled}
#| label: organise_posteriors_controlled
#
# Organising the posteriors of the previous models to fit the tables below
#
source(here("R", "func.R"))
#
# Databases for each region/stimulus
post_OB_cont <- data.frame()
post_OT_cont <- data.frame()
# Names posteriors:
region <- c("OB", "OT")
names_OB_cont <- c("m_controlled_mean_potential_OB", "m_controlled_mean_ros_OB", "m_controlled_mean_dnadamage_OB", "m_controlled_mean_peroxidation_OB")
names_OT_cont <- c("m_controlled_mean_potential_OT", "m_controlled_mean_ros_OT", "m_controlled_mean_dnadamage_OT", "m_controlled_mean_peroxidation_OT")
#
# Organising the results
for (r in region) {
  model_select <- get(paste0("names_", r, "_cont"))   
  for (pos in model_select) {
    model <- get(pos)      # Get the model from the global environment
    post_result <- tidy_post(model)        # Apply tidy_post to each model
    # Add a new column to identify the region and model
    post_result$Region <- r
    post_result$Model <- pos
    # Append to the appropriate data frame
    if (r == "OB") {
      post_OB_cont <- bind_rows(post_OB_cont, post_result)
    } else {
      post_OT_cont <- bind_rows(post_OT_cont, post_result)
    }
  }
}
```

```{r, values_posteriors_controlled}
#| label: values_posteriors_controlled
#
# Extracting the posteriors for the models and the values of interest. Here, I am creating dfs for each of the variables with the values for all the prenatal conditions to make contrasts easier to write.
#
source(here("R", "func.R"))
#
# A) Olfactory bulbs/Chemical stimulus
#
MP_OB_con <- post_values(m_controlled_mean_potential_OB, "none")
ROS_OB_con <- post_values(m_controlled_mean_ros_OB, "none")
DNA_OB_con <- post_values(m_controlled_mean_dnadamage_OB, "none")
LP_OB_con <- post_values(m_controlled_mean_peroxidation_OB, "none")
#
# B) Optic tecta/Visual stimulus
MP_OT_con <- post_values(m_controlled_mean_potential_OT, "none")
ROS_OT_con <- post_values(m_controlled_mean_ros_OT, "none")
DNA_OT_con <- post_values(m_controlled_mean_dnadamage_OT, "sex")
LP_OT_con <- post_values(m_controlled_mean_peroxidation_OT, "none")
```

Table SXX. Estimate of the contrasts between treatments for all the variables analysed, and pMCMC values testing the hypothesis that contrast is different from 0. In bold, the values that are significant at pMCMC < 0.05.
```{r tbl-contrasts_controlled}
#| label: tbl-contrasts_controlled
#| tbl-cap: "Contrasts between prenatal conditions for mitochondrial physiology and learning."
#
source(here("R", "func.R"))
#
# A) Organise df 
var <- c("MP", "ROS", "DNA", "LP")
region <- c("OB", "OT")
#
data_table_controlled <- data.frame()
for(v in var){
  for(r in region){
    x <- paste0(v, "_", r, "_con")
    df <- get(x)
    Temperature <- format_dec(mean(c(df$CORT_Hot, df$Control_Hot)) - mean(c(df$CORT_Cold, df$Control_Cold)), 3)
    pMCMC_temp <- format_p(pmcmc(c(df$CORT_Hot, df$Control_Hot) - c(df$CORT_Cold, df$Control_Cold)), 3, equal = FALSE)
    CORT <- format_dec(mean(c(df$Control_Hot, df$Control_Cold)) - mean(c(df$CORT_Hot, df$CORT_Cold)), 3)
    pMCMC_cort <- format_p(pmcmc(c(df$Control_Hot, df$Control_Cold) - c(df$CORT_Hot, df$CORT_Cold)), 3, equal = FALSE)
    Interaction <- format_dec((mean(df$Control_Hot) - mean(df$CORT_Hot)) - (mean(df$Control_Cold) - mean(df$CORT_Cold)), 3)
    pMCMC_int <- format_p(pmcmc((df$Control_Hot - df$CORT_Hot) - (df$Control_Cold - df$CORT_Cold)), 3, equal = FALSE)
    data_temp <- data.frame(Variable = x,
                          Temperature = as.numeric(Temperature),
                          pMCMC_temp = pMCMC_temp,
                          CORT = as.numeric(CORT),
                          pMCMC_cort = pMCMC_cort,
                          Interaction = as.numeric(Interaction),
                          pMCMC_int = pMCMC_int)
    data_table_controlled <- dplyr::bind_rows(data_table_controlled, data_temp)
  }
}
# Modify the df
data_table_controlled_final <- data_table_controlled %>%
  pivot_longer(cols = c(Temperature, CORT, Interaction), 
               names_to = "Predictor", 
               values_to = "Contrast") %>%
  mutate(
    # Extract the pMCMC values from the corresponding columns
    `pMCMC contrast` = case_when(
      Predictor == "Temperature" ~ pMCMC_temp,
      Predictor == "CORT" ~ pMCMC_cort,
      Predictor == "Interaction" ~ pMCMC_int
    )
  ) %>%
  separate(col = Variable, into = c("Variable", "Region", "control"), sep = "_") %>%
  mutate(
    Variable = case_when(
      Variable == "MP" ~ "Mit potential",
      Variable == "ROS" ~ "ROS",
      Variable == "DNA" ~ "DNA damage",
      Variable == "LP" ~ "Lipid peroxidation",
      Variable == "DET" ~ "Detection latency",
      TRUE ~ Variable
      )
    ) %>%
  mutate(Variable = factor(Variable, levels = c("Mit potential", "ROS", "DNA damage", "Lipid peroxidation"))) %>%
  mutate(
    Region = case_when(
      Region == "OB" ~ "Olfactory bulbs",
      Region == "OT" ~ "Optic tecta",
      TRUE ~ Region
    )
  ) %>%
  dplyr::select(Region, Variable, Predictor, Contrast, `pMCMC contrast`) %>%
  arrange(Region, Variable)
#
# C) Make the contrasts table:
#
set_flextable_defaults(
 font.family = "Times New Roman",
 font.size = 10)
#
contrast_table_controlled <- flextable(data_table_controlled_final) %>%
  align(align = "center", j = c(4,5), part = "body") %>%
  align(align = "center", j = c(1:4), part = "header") %>%
  bold(~`pMCMC contrast` < 0.05, j = c("pMCMC contrast", "Contrast", "Predictor"),
       bold = TRUE) %>%  # Bold when PMCMC is "<0.05"
  bold(~`pMCMC contrast` <0.001, j = c("pMCMC contrast", "Contrast", "Predictor")) %>%  # Bold when PMCMC is "<0.001"
  flextable::compose(i = c(2:12,14:24), j = 1, value = as_paragraph(""), part = "body") %>%
  flextable::compose(i = c(2,3,5,6,8,9,11,12,14,15,17,18,20,21,23,24),
                    j = 2, value = as_paragraph(""), part = "body") %>% # To remove some of the values in the second column
  flextable::hline(i = c(3,6,9,12,15,18,21), j = c(2:5), part = "body") %>% 
  flextable::hline(i = 12, part = "body") %>%
  autofit()
#
contrast_table_controlled
```

Table SXX+1. Summary of the final models for Olfactory Bulbs/Chemical stimulus.
```{r, tbl-results_OB_table_cont}
#| label: results_OB_table_cont
#| tbl-cap: "Results of the models testing for Olfactory Bulbs."
#| tbl-name: "results_OB_cont"
#| tbl-label: "results_OB_cont"
source(here("R", "func.R"))
# 
# A) Refining the df summarizing the posteriors for OB/Chemical stimulus (post_OB_cont)
variable_order <- c("Mit potential",
                     "ROS",
                     "DNA damage",
                     "Peroxidation")
predictor_order <- c("b_Intercept",
                     "b_cortCORT",
                     "b_tempHot",
                     "b_cortCORT:tempHot",
                     "b_age_euthanasia",
                     "b_sexFemale")
post_OB_cont_refined <- post_OB_cont %>%
    mutate(
      Model = case_when(
        Model == "m_controlled_mean_potential_OB" ~ "Mit potential",
        Model == "m_controlled_mean_ros_OB" ~ "ROS",
        Model == "m_controlled_mean_dnadamage_OB" ~ "DNA damage",
        Model == "m_controlled_mean_peroxidation_OB" ~ "Peroxidation",
        Model == "m_controlled_mean_potential_OT" ~ "Mit potential",
        Model == "m_controlled_mean_ros_OT" ~ "ROS",
        Model == "m_controlled_mean_dnadamage_OT" ~ "DNA damage",
        Model == "m_controlled_mean_peroxidation_OT" ~ "Peroxidation",
        TRUE ~ Model
      )
    ) %>%
    mutate(CI_95 = paste0("[", CI_lower, " , ", CI_upper, "]")) %>%
    rename(
      Variable = Model,
      Predictors = Predictor,
      'Estimate Mean' = Mean, 
      '95% CI' = CI_95, 
      PMCMC = PMCMC
    ) %>%
    dplyr::select(
      Variable,
      Predictors,
      'Estimate Mean',
      '95% CI',
      PMCMC) %>%
    mutate(Variable = factor(Variable, levels = variable_order)) %>%
    mutate(Predictors = factor(Predictors, levels = predictor_order)) %>%
    arrange(Variable, Predictors)
#
# B) Create table
## Table format
set_flextable_defaults(
 font.family = "Times New Roman",
 font.size = 10)
#
OB_table_controlled <- flextable(post_OB_cont_refined) %>%
  align(align = "center", j = c(3:5), part = "body") %>%
  align(align = "center", j = c(1:5), part = "header") %>%
  bold(~`PMCMC` < 0.05, j = c("PMCMC", "Estimate Mean", "95% CI", "Predictors"),
       bold = TRUE) %>%  # Bold when PMCMC is "<0.05"
  bold(~`PMCMC` <0.001, j = c("PMCMC", "Estimate Mean", "95% CI", "Predictors")) %>%  # Bold when PMCMC is "<0.001"
  flextable::compose(i = c(2:4,6:8,10:14,16:19), j = 1, value = as_paragraph(""), part = "body") %>% # To remove some of the values in the first column
  autofit()
#
OB_table_controlled
```

```{r, fig-results_OBcont}
#| label: fig-results_OBcont
#| fig-cap: "NEW_OB"
#
source(here("R", "func.R"))
#
# A) Plotting the results for OB/Chemical stimulus using the df from before
plot_potential_OB_con <- plotting(MP_OB_con, "Mit potential")
plot_ROS_OB_con <- plotting(ROS_OB_con, "ROS")
plot_DNA_OB_con <- plotting(DNA_OB_con, "DNA damage")
plot_LP_OB_con <- plotting(LP_OB_con, "Lipid peroxidation")
#
fig_OB_con <- plot_grid(plot_potential_OB_con, plot_ROS_OB_con,
                    plot_DNA_OB_con, plot_LP_OB_con, NULL, NULL,
                    ncol = 2, nrow = 3, rel_heights = c(1, 1, 0.2))
#
# Getting the legend (get_legend() does not work if the legend is on the top or bottom)
plot_legend_bottom_con <- plotting(MP_OB_con, "Mit potential") + theme(legend.position = "bottom", legend.title = element_blank())
gtable_con <- ggplot_gtable(ggplot_build(plot_legend_bottom_con))
legend_plot_con <- gtable_con$grobs[[which(sapply(gtable_con$grobs, function(x) x$name) == "guide-box")]]
# 
# Final composition: Merge everything, adding images and legend
plot_OB_con <- ggdraw(fig_OB_con +
  # Insert legend at the bottom-centered
  draw_grob(legend_plot_con, x = 0.45, y = 0.05, width = 0.0001, height = 0.0001))
# Print final plot
ggsave(here("./output/figures/suppl/new_models/plot_OB_con.png"), plot = plot_OB_con, width = 21, height = 14, units = "cm", dpi = 600, bg = "white")
knitr::include_graphics("./output/figures/suppl/new_models/plot_OB_con.png")
```

Table SXX+2. Summary of the final models for Optic Tecta/Visual stimulus.
```{r, tbl-results_OT_table_cont}
#| label: results_OT_table_cont
#| tbl-cap: "Results of the models testing for Olfactory Bulbs."
#| tbl-name: "results_OT_cont"
#| tbl-label: "results_OT_cont"
source(here("R", "func.R"))
# 
# A) Refining the df summarizing the posteriors for OT/Visual stimulus (post_OT_cont)
variable_order <- c("Mit potential",
                     "ROS",
                     "DNA damage",
                     "Peroxidation")
predictor_order <- c("b_Intercept",
                     "b_cortCORT",
                     "b_tempHot",
                     "b_cortCORT:tempHot",
                     "b_age_euthanasia",
                     "b_sexFemale")
post_OT_cont_refined <- post_OT_cont %>%
    mutate(
      Model = case_when(
        Model == "m_controlled_mean_potential_OB" ~ "Mit potential",
        Model == "m_controlled_mean_ros_OB" ~ "ROS",
        Model == "m_controlled_mean_dnadamage_OB" ~ "DNA damage",
        Model == "m_controlled_mean_peroxidation_OB" ~ "Peroxidation",
        Model == "m_controlled_mean_potential_OT" ~ "Mit potential",
        Model == "m_controlled_mean_ros_OT" ~ "ROS",
        Model == "m_controlled_mean_dnadamage_OT" ~ "DNA damage",
        Model == "m_controlled_mean_peroxidation_OT" ~ "Peroxidation",
        TRUE ~ Model
      )
    ) %>%
    mutate(CI_95 = paste0("[", CI_lower, " , ", CI_upper, "]")) %>%
    rename(
      Variable = Model,
      Predictors = Predictor,
      'Estimate Mean' = Mean, 
      '95% CI' = CI_95, 
      PMCMC = PMCMC
    ) %>%
    dplyr::select(
      Variable,
      Predictors,
      'Estimate Mean',
      '95% CI',
      PMCMC) %>%
    mutate(Variable = factor(Variable, levels = variable_order)) %>%
    mutate(Predictors = factor(Predictors, levels = predictor_order)) %>%
    arrange(Variable, Predictors)
#
# B) Create table
## Table format
set_flextable_defaults(
 font.family = "Times New Roman",
 font.size = 10)
#
OT_table_controlled <- flextable(post_OT_cont_refined) %>%
  align(align = "center", j = c(3:5), part = "body") %>%
  align(align = "center", j = c(1:5), part = "header") %>%
  bold(~`PMCMC` < 0.05, j = c("PMCMC", "Estimate Mean", "95% CI", "Predictors"),
       bold = TRUE) %>%  # Bold when PMCMC is "<0.05"
  bold(~`PMCMC` <0.001, j = c("PMCMC", "Estimate Mean", "95% CI", "Predictors")) %>%  # Bold when PMCMC is "<0.001"
  flextable::compose(i = c(2:4,6:8,10:13,15:18), j = 1, value = as_paragraph(""), part = "body") %>% # To remove some of the values in the first column
  autofit()
#
OT_table_controlled
```

```{r, fig-results_OTcont}
#| label: fig-results_OTcont
#| fig-cap: "NEW_OT"
#
source(here("R", "func.R"))
#
# A) Plotting the results for OT/Visual stimulus using the df from before
plot_potential_OT_con <- plotting(MP_OT_con, "Mit potential")
plot_ROS_OT_con <- plotting(ROS_OT_con, "ROS")
plot_DNA_OT_con <- plotting(DNA_OT_con, "DNA damage")
plot_LP_OT_con <- plotting(LP_OT_con, "Lipid peroxidation")
#
fig_OT_con <- plot_grid(plot_potential_OT_con, plot_ROS_OT_con,
                    plot_DNA_OT_con, plot_LP_OT_con, NULL, NULL,
                    ncol = 2, nrow = 3, rel_heights = c(1, 1, 0.2))
#
# Getting the legend (get_legend() does not work if the legend is on the top or bottom)
plot_legend_bottom_con <- plotting(MP_OT_con, "Mit potential") + theme(legend.position = "bottom", legend.title = element_blank())
gtable_con <- ggplot_gtable(ggplot_build(plot_legend_bottom_con))
legend_plot_con <- gtable_con$grobs[[which(sapply(gtable_con$grobs, function(x) x$name) == "guide-box")]]
# 
# Final composition: Merge everything, adding images and legend
plot_OT_con <- ggdraw(fig_OT_con +
  # Insert legend at the bottom-centered
  draw_grob(legend_plot_con, x = 0.45, y = 0.05, width = 0.0001, height = 0.0001))
# Print final plot
ggsave(here("./output/figures/suppl/new_models/plot_OT_con.png"), plot = plot_OT_con, width = 21, height = 14, units = "cm", dpi = 600, bg = "white")
knitr::include_graphics("./output/figures/suppl/new_models/plot_OT_con.png")
```
